### 2020-11-18 docs/manuscript.tex
add 'System and Methods' section

\PaperTitle{
    HAMDetector: Combining information to detect HLA-associated mutations with a
    Bayesian regression model
} % Article title

@@

  \textbf{Availability and Implementation} \\
  The source code of this software is available at 
  \url{http://gogs.uni-due.de/habermann/Escape.git} under a permissive MIT license. \\

@@

% - subset of peptidome presented on the cell surface via MHC I.
% - Cytotoxic t cells are selected to not recognize self.
% - by rule of exclusion, recognized peptides are foreign
% - t cells induce cytolytic activity and recruit other immune cells, reference

% - basic function of statistical analysis of sequence data
% - state of the art + brumme paper, phylogeny, linkage disequilibrium and codon covariation
% cite brumme and other paper

% - problems of p value based methods and statistical significance testing
% - highlight benefits of Bayesian approaches
\subsection{The HLA system}

@@

Upon transmission, HLA escape mutations typically quickly revert to their wild type
as they usually reduce viral replicative capacity (if a mutation would increase
viral replicative capacity regardless of the presence of a given HLA allele, it would
probably already be the wild type). Kawashima et al.
describe an escape mutation that is selected by HLA allele B\*53,
does not strongly affect viral replicative capacity and therefore slowly enriches
over time in Japan, where B\*53 commonly occurs.
How quickly a given escape mutation is selected upon transmission in a host depends on the
magnitude of the reduction in viral replicative capacity, on the strength of selection
pressure and also on the genetic background, e.g. some escape mutations require compensatory 
mutations which partly attenuate the negative impact on viral replicative capacity.

@@

Fisher\'s exact test has the advantage of being easy to apply, but has also several 
disadvantages that are outlined in Carlson et al.
The most striking one is that viral sequences share a common phylogenetic history and therfore,
treating sequences as independent and identically distributed samples may under- or
overestimate effect sizes and in the context of hypothesis testing leads to
increased false positive and false negative rates.
Another issue of applying Fisher\'s exact test is that HLA class I loci are located in
close proximity on chromosome 6 and are therefore in linkage disquilibrium, which means 
that HLA alleles are not inherited completely independent of each other, so that
inheritance of one HLA allele correlates with inheritance of another HLA allele.
When using a statistical method that tests each HLA allele individually without
considering the whole set of alleles present, spurious associations might occur
: If HLA allele \(H_{1}\) is associated with an amino acid replacement \(R\), but
\(H_{1}\) is in linkage disequilibrium with another allele \(H_{2}\), this also means that
we observe an association between replacement \(R\) and \(H_{2}\), even in
absence of any underlying escape mechanism.

@@

Another issue that occurs when using p-values as a screening tool is the problem
of multiple comparisons. When applying a statistical test, the probability  of 
obtaining a statistically significant result increases with
each additional test, even in absence of any real effect. When using p-values as a filter
it is therefore likely to obtain significant effects that are in fact not real.
To circumvent this problem, a common strategy is to control the false discovery
rate, which is the expected proportion of false positives.
These adjustment procedures have the problem that, when performing many of such comparisons,
none but the very largest effects remain.
Instead of performing many hypothesis tests and trying to adjust for them, we instead prefer
to fit a single, multilevel model that contains all comparisons of interest.
In this setting, the problem of multiple comparisons can disappear entirely and also
yield more valid estimates.

@@

>>>>

\PaperTitle{
    HAMdetector: Combining information to detect HLA-associated mutations with a
    Bayesian regression model
} % Article title

@@

  \textbf{Availability and Implementation} \\
  The source code of this software is available at 
  \url{http://github.com} under a permissive MIT license. \\

@@

\subsection{The HLA system}

@@

Upon transmission, HLA escape mutations typically quickly revert to their wild type
as they usually reduce viral replicative capacity (if a mutation would increase
viral replicative capacity regardless of the presence of a given HLA allele, it would
probably already be the wild type). Kawashima et al.
describe an escape mutation that is selected by HLA allele B*51,
does not strongly affect viral replicative capacity and therefore slowly enriches
over time in Japan, where B*53 commonly occurs.
How quickly a given escape mutation is selected upon transmission in a host depends on the
magnitude of the reduction in viral replicative capacity, on the strength of selection
pressure and also on the genetic background, e.g. some escape mutations require compensatory 
mutations which partly attenuate the negative impact on viral replicative capacity.

@@

Fisher's exact test has the advantage of being easy to apply, but has also several 
disadvantages that are outlined in Carlson et al.
The most striking one is that viral sequences share a common phylogenetic history and therfore,
treating sequences as independent and identically distributed samples may under- or
overestimate effect sizes and in the context of hypothesis testing leads to
increased false positive and false negative rates.
Another issue of applying Fisher's exact test is that HLA class I loci are located in
close proximity on chromosome 6 and are therefore in linkage disquilibrium, which means 
that HLA alleles are not inherited completely independent of each other, so that
inheritance of one HLA allele correlates with inheritance of another HLA allele.
When using a statistical method that tests each HLA allele individually without
considering the whole set of alleles present, spurious associations might occur
: If HLA allele \(H_{1}\) is associated with an amino acid replacement \(R\), but
\(H_{1}\) is in linkage disequilibrium with another allele \(H_{2}\), this also means that
we observe an association between replacement \(R\) and \(H_{2}\), even in
absence of any underlying escape mechanism.

@@

Another issue that occurs when using p-values as a screening tool is the problem
of multiple comparisons. When applying a statistical test, the probability  of 
obtaining a statistically significant result increases with
each additional test, even in absence of any real effect. When using p-values as a filter
it is therefore likely to obtain significant effects that are in fact not real.
To circumvent this problem, a common strategy is to control the false discovery
rate, which is the expected proportion of false positives.
These adjustment procedures have the problem that, when performing many of such comparisons,
none but the very largest effects remain.
Instead of performing many hypothesis tests and trying to adjust for them we instead prefer
to fit a single, multilevel model that contains all comparisons of interest.
When using multilevel models, the problem of multiple comparisons can disappear entirely and 
also yield more valid estimates.

@@

When possible, we choose to fit Bayesian models. By using prior information,
adding problem-specific structure and partial pooling, the accuracy of estimates
can often be noticably improved. Prior information does not necessarily mean to
use external data, even a rough idea about the expected magnitude of estimates is often
surprisingly effective.
Additionally, Bayesian statistics provides an accessible way to test models:
By comparing data generated under the model's assumptions to the actually observed data,
it is possible to identify important aspects of the dataset that the models fails to
capture and subsequently improve the model until it is consistent with the observed data.

In the context of identifying HLA-associated mutations we propose to improve existing
methods by the following additions, which can be broadly divided into additional information
and model structure:
\subsection*{Additional information}

  \textit{Binding affinity prediction}

    HLA-associated mutations are expected to lie more frequently in regions of known
    epitopes. There are vast experimental binding affinity data available of different
    peptides and MHC I molecule pairs and there are well-established computational
    methods that use these data to extrapolate HLA binding for untested peptides.
    We show that by including the outcome of these computational tools as input for a
    probabilistic model, the prediction of HLA-associated mutations can be improved.

  \textit{Antigen processing prediction}

  Similarly, there are also HLA-allele independent effects like antigen processing
  that influence presentation on MHC I. Second generational tools like
  MHCFlurry 2.0 use both binding affinity and antigen processing data to improve
  epitope prediction. For our tool HAMdetector, we use the output of MHCFlurry to
  benefit from this binding affinity and antigen processing data for predicting
  HLA-associated mutations.

\subsection*{Model structure}

  \textit{Sparsity-inducing priors}

  Recent advantages in Bayesian inference include so-called sparsity-promoting priors.
  Sparsity-promoting priors convey the apriori expectation that most coefficients in 
  a regression model are close to 0.
  This assumption also applies to HLA-associated mutations, because the number of epitopes
  that are restricted by a given HLA allele is typically small compared to the number
  of all possible epitopes. If no epitope spanning a given alignment position is presented
  on the MHC I molecule, any association between a replacement and the respective HLA allele 
  is likely to be due to random variation alone (or HLA linkage disequilibrium).

  It has been shown that using a sparsity-promoting prior when non-zero coefficients
  are sparse can drastically improve predictive performance because the model is
  better able to differentiate between signal and noise.

  We show that a simple logistic regression model with a sparsity-promoting prior
  on the regression coefficients for the HLA alleles alone already performs roughly on-par 
  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et. al.
  when using the fraction of identified escapes that lie in known HLA epitopes as a
  benchmark.

  \textit{Partial pooling of 4-digit HLA alleles}

  It has recently been appreciated that binding specifities can vary drastically across 
  HLA alleles from the same allele group, e.g. between HLA-B*51:01 and HLA-B*51:03.
  For predicting HLA associated mutations, there has therefore been a shift to use
  4-digit resolution data whenever possible.
  This is not without downsides however, because overall, binding specifities
  are more similar for alleles in the same group and therefore, treating them
  as completely separate might unnecessarily fragment the available data.

  In Bayesian statistics, it is not required to chose one extreme or the other:
  Instead of either treating all HLA alleles from the same allele group as identical or
  as completely separate, partial pooling of the estimates allows something inbetween:
  By partial pooling, estimates do share some information across each other, but they are
  also allowed to vary if necessary.

  In the context of our model this means that observing a relevant association between
  a replacement and an HLA allele also makes the model more inclined to estimate
  relevant associations between that replacement and all other alleles from that allele group.
  The degree of this partial pooling can be estimated from the data, so if we do observe
  strong similarity of HLA alleles in a given allele group, the estimates are more
  influenced by each other than if they do not.
